---
title: "`wbt()` method: call 'WhiteboxTools' by name"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`wbt()` and `wbt_result`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = whitebox::check_whitebox_binary() & 
          requireNamespace("terra", quietly = TRUE)
)
```

# `wbt()`

The `wbt()` method is a high-level wrapper function to manage flow of
information between **R** and 'WhiteboxTools'. It allows you to call 'WhiteboxTools' by name, and returns a result object that provides information about the execution of the specified tool.

When a tool is executed using the `wbt()` method, the result object contains information about the execution of the tool, including the parameters that were passed to the 'WhiteboxTools' executable and references to any output file paths that were specified. If the system call succeeds, the result object provides information about the output, such as the data type and file path, which can be used for further analysis or integration with other spatial data sources and R object types.

In the case of a failed system call, the result object may contain an error message. This provides valuable feedback to users for troubleshooting and debugging their code. By using the `wbt()` method, users can efficiently integrate 'WhiteboxTools' algorithms into their R workflows and take advantage of the powerful spatial analysis capabilities of 'WhiteboxTools' within a familiar R environment.

This vignette provides several examples of using the `wbt()` method to call 'WhiteboxTools' by name with various R objects as input and illustrates how to work with the resulting output and result objects in R.

```{r}
library(whitebox)
```

## Getting Started

Know what tool you want to run but _can't remember the required parameters_?
Just type `wbt("toolname")`

```{r}
wbt("slope")
```

Output produced on error and with invalid arguments helps guide use
interactively by prompting with required and optional arguments as a checklist.

## Result Objects

All calls to `wbt()` return an S3 object with class `wbt_result`.

```{r}
# get file path of sample DEM
dem <- sample_dem_data()

wbt("slope", dem = dem, output = file.path(tempdir(), "slope.tif"))
```

If the output files specified by the tool reference spatial data, such as a GeoTIFF file, the `wbt()` method can convert them into R spatial objects. These file-based R objects can be further processed, analyzed, and visualized using various R packages such as terra, raster, or stars. The result object also provides information about the output R objects, such as the data type and file path, which can be used for further analysis or integration with other spatial data sources.

```{r}
wbt("slope", goof = dem, output = file.path(tempdir(), "slope.tif"))
```

When a call fails an informative error message issued, and the error will be cached
inside the result. Prior runs references are stored as well for sequential tool runs; more on that below.

### `wbt_result` class

A `wbt_result` class object is a base R `list()` with standard element
names (`"tool"`, `"args"`, `"stdout"`, `"crs"`, `"result"`, `"history"`) and
the `"class"` attribute `"wbt_result"`.

```{r}
str(wbt("slope", dem = dem, output = file.path(tempdir(), "slope.tif")), max.level = 1)
```

Any `output` produced by the tool (usually file paths set by the user) will be
included in the `wbt_result$result` list.

### `wbt_result` on error

If there is an error a `try-error` class object with an error message is
returned in lieu of a list in `$result`

```{r}
# on error there is a try-error object in $result
x <- wbt("slope")
inherits(x$result, 'try-error')
message(x$result[1])
```

## Vignette Topics

We now will cover how the `wbt()` method and `wbt_result` class can be used for
the following:

 - Input and Output with R spatial objects

 - Running Sequences of Tools (optionally: with pipe `|>` or `%>%`)

 - Handling Coordinate Reference Systems

# Input and Output with R spatial objects

A feature of `wbt()` is that it handles input/output and file name management
if you are using R objects as input. It can be an onerous task to manage files 
for workflows involving many tool runs.

If you use a `terra` object as input object, you will get a `SpatRaster`. If you use 
a `raster` object as your input, you will get a `RasterLayer`
object as output for tools that produce a raster. There will be file paths 
associated with that R object. 

### Compare `raster` and `terra` as R raster object frontends

```{r}
if (requireNamespace("raster")) {
  rdem <- raster::raster(dem)

  # raster input; raster output
  r1 <- wbt("slope", dem = rdem, output = file.path(tempdir(), "slope.tif"))
  r1
  class(r1$result$output)
}
```

```{r}
tdem <- terra::rast(dem)

## terra input; terra output
t1 <- wbt("slope", dem = tdem, output =  file.path(tempdir(), "slope.tif"))
t1
class(t1$result$output)
```

The user still needs to specify paths but `wbt()` eases the process
of converting input objects to paths and reading output paths to R objects
after the tool has run. In principle any R object that references a file with
type that 'WhiteboxTools' supports can be handled under the current (simple; proof of
concept) system. Using file-based representations may not be the most
efficient when your R object only previously existed in memory, it allows 
for a tangible file "trail" of steps in a workflow that were followed that 
otherwise might be hard to track.

# `wbt_source()`: rasters in memory, and vector data

The terra and raster packages typically use pointers to a file/raster connection, except when a grid is entirely in memory. When processing in-memory rasters with 'WhiteboxTools', it's necessary to write out the object as a temporary GeoTIFF file for tool input. Fortunately, the `wbt()` method handles this step automatically.

Vector data in the form of point, line, and polygon geometries can be stored in sf data.frames, sp objects, or terra SpatVector objects. These objects are read into memory when created in R. As it's sometimes necessary to process vector data with 'WhiteboxTools', the `wbt_source()` method can read vector data and annotate it with information about the source file path. By using `wbt_source()`, users can process vector data with `wbt()` while minimizing overhead, i.e., without creating a temporary file for 'WhiteboxTools' to read.

The `wbt_source()` method can create a reference for a sample ESRI Shapefile or any other vector data source that can be read by sf, sp, or terra. The result is a SpatVectorProxy object that can be used with `wbt()` for processing spatial data. If a file path is passed to `wbt_source()`, a SpatVectorProxy with the necessary information is created automatically. For instance, to create a reference for a sample ESRI Shapefile, users can call `wbt_source()` and pass the file path as an argument:

```{r}
library(terra)
shp <- system.file("ex/lux.shp", package = "terra")
x <- wbt_source(shp)
x
```

The result is a SpatVectorProxy.

The SpatVectorProxy provides a light-weight means to create an R object reference to a large file; allowing the "heavy" lifting for that file to be done primarily by WhiteboxTools (or GDAL via {terra}).

At time of writing this document, WhiteboxTools does not support vector data inputs other than Shapefile. If we have, for example, a GeoPackage, `wbt_source()` will convert the input to a temporary shapefile and return a SpatVectorProxy that references that file. This object can then be used as input to `wbt()`.

```{r}
# load the data
x2 <- query(x)

# remove area column
x2$AREA <- NULL

# create a GeoPackage
terra::writeVector(x2, filename = file.path(tempdir(), "lux.gpkg"), overwrite = TRUE)

# now the source is a temporary .shp
x3 <- wbt_source(file.path(tempdir(), "lux.gpkg"))

wbt("polygon_area", input = x3)
```

# Running Sequences of Tools

When you call `wbt()` and the first argument is a `wbt_result` the `output`
from the input `wbt_result` is passed as the first `input` to the new tool run.

This general "pass the output as the first input" works even if the first
`input` parameter (`"-i"` flag) is something different such as `dem` or
`input1`.

This makes it possible to chain different operations together in a sequence.

If all of the required parameters are specified the tool will be run.

Here we run "Slope" on a DEM, then run "Slope" on it again to get a second
derivative "curvature".

```{r}
x <- wbt("slope", dem = dem, output = file.path(tempdir(), "slope.tif"))
x2 <- wbt(x, tool_name = "slope", output = file.path(tempdir(), "curvature.tif"))
```

## Using pipes

Users can use pipe operators to transform sequences of 'WhiteboxTools' calls that depend on each other. In magrittr's `%>%` and base R's `|>` syntax (R 4.1+), the output of the function on the left-hand side is automatically set as the first argument of the function on the right-hand side. This allows users to chain functions together, passing the output of one function as the input to the next, and can significantly simplify the code and improve readability.

```{r, eval = FALSE, purl = FALSE}
wbt("slope", dem = dem, output = file.path(tempdir(), "slope.tif")) |>
  wbt("slope", output = file.path(tempdir(), "curvature.tif"))
```

The `wbt_result` output that prints to the console will reflect the
input/output parameters of the most recent tool run in a tool chain.

```{r}
x2
```

`wbt_result()` is the method to get `$result` for single runs or all `$result`
in `$history` for chained runs from a `wbt_result` object.

```{r}
str(wbt_result(x2), max.level = 1)
```

The `$history` element is a list of the `wbt_result` from individual runs.

```{r}
str(x2$history, max.level = 1)
```

If you pass invalid results from one step to the next, you get an informative
error message.

```{r}
x <- wbt("slope")
wbt(x, "slope", output = file.path(tempdir(), "foo.tif"))
```

# Coordinate Reference Systems (CRS)

On top of handling file paths it is necessary for Geographic Information
Systems to handle Coordinate Reference System (CRS) information.  Many R tools
in the GDAL/PROJ sphere require valid CRS to operate.

WhiteboxTools will read the GeoKey information from your _GeoTIFF_ files
specified by path and propagate that to the output file result. When you read
those files with R, you should find that the original file's CRS has been
transferred.

`wbt()` and `wbt_result` help ensure consistent file and CRS information across
sequences of operations and within calls--especially those _involving R spatial
objects_.

If you specified the CRS in R, or made the raster entirely in R, you might need
a hand setting up the file-based representation that WhiteboxTools will use.
This is easy enough to do with `crs()`, `writeRaster()` and the equivalent but
often requires more boilerplate code than just specifying the argument, or
having CRS propagate from input.

For inputs that have a file-based representation (e.g. `RasterLayer` and
`SpatRaster` objects) `wbt()` provides an interface where R objects can be
substituted for file names and vice versa. User and workflow level attributes
like CRS and working directory be handled at the same time.

```{r}
dem <- sample_dem_data()

## equivalent to:
# dem <- system.file("extdata/DEM.tif", package = "whitebox")
```

In the process of reading/writing files that may contain CRS information, such
as this sample DEM, the CRS can be inspected, modified and propagated to file
or R objects.

## Setting the CRS

If the CRS is specified in the input `SpatRaster` object, `wbt()` makes sure it is
propagated to the R object result. `wbt()` does not alter or copy the source or 
output files which may or may not have CRS information.

```{r}
araster <- terra::rast(dem)
wbt("slope", dem = araster, output = file.path(tempdir(), "foo.tif"))
```

Alternately you can specify the `crs` argument to `wbt()` instead of `rast()`. This will directly set the `crs` element of the `wbt_result` of your output.

```{r}
wbt("slope", dem = terra::rast(dem), crs = "EPSG:26918", output = file.path(tempdir(), "foo.tif"))
```

In either case wherever you specify the `crs` argument the `crs` is
stored in the `wbt_result` and propagated to the relevant output. Again, the
source files are unchanged if they had no CRS or invalid CRS (until you write that updated CRS to file).

If two input R objects have different (or no) CRS then the `wbt()` method will 
provide a message to the user:

```{r}
r1 <- terra::rast(dem) # default: EPSG:26918
r2 <- terra::deepcopy(r1)
crs(r2) <- "EPSG:26917" # something else/wrong
wbt("add", 
    input1 = r1, 
    input2 = r2, 
    output = file.path(tempdir(), "foo.tif")
   )
```

This message is suppressed and output CRS set accordingly if the `crs` argument is specified. This user set CRS is only in the SpatRaster object, and does not necessarily match that of the file used as input or returned as output.

```{r}
wbt("add", 
    input1 = r1,
    input2 = r2, 
    crs = "EPSG:26918",
    output = file.path(tempdir(), "foo.tif")
   )
```
 <!--
# Future Plans

Future additions to this vignette may include:

 - Managing WhiteboxTools working directories

 - Managing WhiteboxTools file types
 
In the future users will be able to:

 - Set a temporary or persistent working
directory and have output file paths/names auto-generated, allowing them to
work entirely with R objects and `wbt_result` while making an intermediate file
record in the background.

 - Use a wrapper class for generic
groups of Whitebox file types; allowing file name references for any R object
or output file type to be stored: optimized/customizable read/write methods used for
specific file types and generic .rda files with metadata for others.

 - Batch of tool runs on e.g. watershed extents -->
 

```{r, echo=FALSE} 
# cleanup temp files
unlink(list.files(".", pattern = "file.*tif$", full.names = TRUE))
unlink(list.files(tempdir(), pattern = "file.*tif$", full.names = TRUE))
```
